---
title: "Walsh_Figueroa_Data_Cleaning"
author: "Wes Walsh"
date: "2025-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up
```{r Load packages, include=FALSE}
#Get packages used
require(here)
library(tidyverse)
library(rgbif)

#citation()
#version$version.string
```

NatureServe Full Canada and US download
Report Generated October 16, 2025
Search Criteria: 
Include only full species.
Include only species with accepted taxonomy.
Location: Canada, United States
Origin: Native
Scientific Taxonomy: CLASS Arachnida (Animalia),CLASS Insecta (Animalia)

#Get Data
```{r Get data}

na_full <- read.csv(here("data", "natureserve_NA_noexotic_full.csv"))

#Remove blank entries
na_full <- na_full[-which(na_full$scientificName == ""), ]

#Convert to factor
na_full <- na_full |>
  mutate(across(c(Global_Ranked, Global_Rank_Rnd), as.factor))


```

#Match to GBIF Backbone
```{r GBIF match}

#Just scientific name column
scientificName <- na_full["scientificName"]

#Match names against GBIF backbone
name_match <- name_backbone_checklist(scientificName)

#Just taxonomy
taxonomy <- name_match %>%
      dplyr::select(verbatim_name,synonym,class,order,family,genus) %>%
      rename(scientificName = verbatim_name)

#Merge based on name
scientificName <- merge(scientificName, taxonomy, by='scientificName', all.y=TRUE)

```

#Format NatureServe
```{r Format NatureServe}
#Merge based on name
na_full <- merge(scientificName,na_full, by = 'scientificName', all.x=TRUE)


#Just columns I want
na_full <- na_full %>%
      dplyr::select(scientificName, synonym, class, order, family, genus, commonName, Synonyms, Species.Group..Fine., Global_Ranked, Global_Rank_Rnd, U.S..Endangered.Species.Act.Status, Distribution, View.on.NatureServe.Explorer)

#Save as csv file
write.csv(na_full, "NA_full_taxonomy.csv")

```

NatureServe US Download
Report Generated October 16, 2025
Search Criteria: 
Include full species, subspecies, varieties, and populations.
Include only species with accepted taxonomy.
Location: United States
Origin: Native
Scientific Taxonomy: CLASS Arachnida (Animalia),CLASS Insecta (Animalia)

#Get Data
```{r Get data}

us_full <- read.csv(here("data", "us_full_noexotic.csv"))

#Rename column
us_full <- us_full %>%
  rename(scientificName = Scientific.Name)

#Remove blank entries
us_full <- us_full[-which(us_full$scientificName == ""), ]

```

#Match to GBIF Backbone
```{r GBIF match}

#Just scientific name column
us_scientificName <- us_full["scientificName"]

#Match names against GBIF backbone
name_match <- name_backbone_checklist(us_scientificName)

#Just taxonomy
taxonomy <- name_match %>%
      dplyr::select(verbatim_name,synonym,class,order,family,genus) %>%
      rename(scientificName = verbatim_name)

#Merge based on name
us_scientificName <- merge(us_scientificName, taxonomy, by='scientificName', all.y=TRUE)

```

#Format NatureServe
```{r Format NatureServe}
#Merge based on name
us_full <- merge(us_scientificName,us_full, by = 'scientificName', all.x=TRUE)


#Just columns I want
us_full <- us_full %>%
      dplyr::select(scientificName, synonym, class, order, family, genus, Common.Name, Synonyms, Species.Group..Fine., NatureServe.Global.Rank, NatureServe.Rounded.Global.Rank, U.S..Endangered.Species.Act.Status, Distribution, View.on.NatureServe.Explorer)

#Save as csv file
write.csv(us_full, "us_full_taxonomy.csv")

```

BIRDS

#Get Data
```{r Get data}

us_fullbirds <- read.csv(here("data", "us_fullbirds_noexotic.csv"))

```

#Match to GBIF Backbone
```{r GBIF match}

#Just scientific name column
usbirds_scientificName <- us_fullbirds["scientificName"]

#Match names against GBIF backbone
name_match <- name_backbone_checklist(usbirds_scientificName)

#Just taxonomy
taxonomy <- name_match %>%
      dplyr::select(verbatim_name,class,order,family,genus) %>%
      rename(scientificName = verbatim_name)

#Merge based on name
usbirds_scientificName <- merge(usbirds_scientificName, taxonomy, by='scientificName', all.y=TRUE)

```

#Format NatureServe
```{r Format NatureServe}
#Merge based on name
us_fullbirds <- merge(usbirds_scientificName,us_fullbirds, by = 'scientificName', all.x=TRUE)

#Save as csv file
write.csv(us_fullbirds, "us_fullbirds_taxonomy.csv")

```